# `R` Objects 

`R` is an **object-oriented** programming language, meaning we will almost always store data in **objects** and run **functions** on those objects. We **assign** values to objects using the **assignment operator (`<-`)**[^1] 
```{r, eval=F}
myobject<-value
```

The keyboard shortcut for inserting `<-` (inside the Console or an `R` chunk) is `Alt`+`-` (on Windows) or `Option`+`-` (on Mac).


[^1]: Think of the assignment operator like like an $=$ sign, but we want to avoid using the equals sign. `<-` was [originally its own key](https://twitter.com/kwbroman/status/747829864091127809) on early computer keyboards. 

**Functions** take the form: 
```{r, eval=F}
functionname(myobject)
```

Functions can have *other* functions for **arguments** (the object the function is run on), e.g. 
            
```{r, eval=F}
round(rnorm(5),2) 
# rnorm(5) takes 5 random draws from a normal distribution
# then round(, 2) rounds the result to 2 decimal places
```

## Vectors

The simplest data structure in `R` is a **vector**, simply a collection of objects or elements. To construct a vector, use the "combine/concatenate" function "`c()`"

As an example, let's make a vector of the numbers 1 through 5, called `v`. 

```{r}
v<-c(1,2,3,4,5)
```

We can also build vectors via generating mathematical **series** with the `:` operator, which lists all integers in a series from `beginning:end`. 

```{r}
v<-1:5
```

To inspect an object, we simply "call" it up by typing the name of the object to print its contents.

```{r}
v
```

### Functions

Since a vector is an object, we can run functions on that object. Let's start with some simple mathematical functions, such as taking the sum and taking the mean of our simple vector `v`. 

```{r}
sum(v)
mean(v)
```



Functions in `R` are "vectorized," meaning the function is run on every object inside a vector. 

We can perform mathematical operations on a vector as a whole:

```{r}
sum(1:5)
mean(1:5)
```

## Other Object Types

### Lists 

- A **list** is a non-atomic vector, meaning you can gather data elements of different classes in one object 

```{r list}
mylist<-list(5, pi, TRUE, 4.3, "cabbage")
class(mylist)
```

- Another great property of lists is that elements of the list can themselves be vectors

```{r}
vectored.list<-list(c(1.82, 1940, 93.20, 192.917),
     c("Orange", "Cyan", "Pink"),
     c(TRUE, FALSE, TRUE, TRUE, FALSE, TRUE, FALSE, FALSE))

str(vectored.list) # look at structure of the list
```

- We can create a **label** for each element in a list, called a `name` 

```{r}
vectored.list<-list(numbers=c(1.82, 1940, 93.20, 192.917), # first element is a vector called 'numbers'
     colors=c("Orange", "Cyan", "Pink"), # second element is a vector called `colors`
     logic=c(TRUE, FALSE, TRUE, TRUE, FALSE, TRUE, FALSE, FALSE)) # third element is a vector called `logic`

vectored.list
```

- The `names` command prints (or changes) the name of the label of each element in the list

```{r}
names(vectored.list) # print the names of the list elements
names(vectored.list)<-c("name1","name2","name3") # rename the lables to 'name1', 'name2', and 'name3'

names(vectored.list) # print new names
vectored.list # print list with new names
```

### Matrix

- Everything thus far has been 1 dimension, but we often work with 2-dimensional data
    - Rows are observations
    - Columns are variables
- A **matrix**
    - `matrix()` command creates a matrix by column,
        - can define number of rows with `nrow=`, `R` will divide the elements into equal number of columns 
    
```{r}
matrix1<-matrix(c(1,2,3,4,5,6),nrow=3) # make a 3-row matrix
matrix1 
```

### Data Frame

- The most important object in `R` is a **data frame** (what you call a "spreadsheet"), used for statistics, plots, regressions, etc
    - "Rectangular" data, rows are observations, columns are variables 
    - Can hold variables of different classes (e.g. a quantitative variable like income, a character variable like name, etc)
    - In essence, data frames are actually lists (where each list object itself is a vector)
    - All vectors (columns) must have the same length! 

```{r}

df<-data.frame(x = 1:3,
           y = c("a", "b", "c"),
           z = c(TRUE, FALSE, TRUE))
df
```

## Data Classes


- Vectors **must** contain the same type of elements (e.g. numerical or text)
  * Technically this refers to **atomic vectors** (nearly all vectors are atomic)
- Vectors with "mixed" types will convert all elements to the lowest-common denominator, e.g. character
- You can always check the type of vector using **class()**
```{r vector.types}
mixed<-c(5, pi, TRUE, 4.3, "cabbage")
class(mixed)
```

### Numeric

- **Numeric** (aka "double"), as it sounds, can perform mathematical operations
```{r numeric, echo=TRUE, eval=FALSE}
numeric<-c(1,2,3,4,5)
```

- There are two **types** of numeric objects: **double** and **integer** 

### Double

- If numeric values contain decimal points, they are technically called **floating point double** or simply **double** class
- `R` may simply call them numeric, but contrast with integer below 

```{r}
double<-c(pi,2.34,9.99)

class(double)
typeof(double) # will return the more specific type 

is.double(double) # a logical test to see if object is "double" type
is.integer(double) # a logical test to see if object is "integer" type
```

#### Integer

- If numeric values are all whole numbers, they are **integer** class

```{r}
integers<-c(1,2,3,4)
class(integers)
typeof(integers)

is.double(double)
is.integer(double)
```

### Logical

- **Logical** is a series of binary elements or statements that can either be TRUE or FALSE

```{r logical, echo=TRUE,eval=FALSE}
logical<-c(TRUE,FALSE,FALSE,TRUE)
```

- We can perform logical tests with common operators:
    - `<` less than
    - `>` greater than
    - `<=` less than or equal to ($\leq$)
    - `>=` greater than or equal to ($\geq$)
    - `==` is equal to (note two equals signs are needed!)
    - `!=` is not equal to 
    - `%in%` is a member of a set ($\in$)

```{r}
3==4 #is 3 equal to 4?
3<4 # is three less than 4?
3<=4 # is three less than or equal to 4?
3>4 # is three greater than 4?
3!=4 # is three not equal to four?
3 %in% c(0,1,2) # is three in the following set of numbers?
3 %in% c(0,1,2,3) # is three in the following set of numbers?
```

- We are not limited to using numeric data, `R` can also perform logical tests on other classes of variable, like characters (which need quotes): 

```{r}
"red"=="blue" # is red the same as blue?
"red"!="blue" # is red not equal to blue?
political.party<-c("Republican","Democrat") # define political party as a set of Republican and Democrat
"Libertarian" %in% political.party # check if Libertarian is in the set of political parties we created
"Democrat" %in% political.party # check if Democrat is in our set of political parties 
```

- We can also perform more than one test at a time with multiple conditions:
    - `&` AND 
    - `|` OR 

```{r}
2==2 & 2>3 # is 2 equal to 2 AND greater than 3? 
2==2 | 2>3 # is 2 equal to 2 OR greater than 3? 
```

- These commands will become very useful when we want to subset data or look at portions of our data based on some condition

### Character

- **Character** is a string of text: letters, numbers, and symbols, cannot perform mathematical operations
    - Character values require quotation marks around each value 
```{r char, echo=TRUE,eval=FALSE}
character<-c("one","two","7","orange")
```

#### Dates

- Dates are a specific type of character class
- Specific dates
    * Can do days, weeks, months, quarters, years
```{r dates 1}
today<-Sys.Date() #print today's date 
format(today, format="%B %d %Y") # specify how to report date format 

months<-seq(as.Date("2010/1/1"), as.Date("2012/1/1"), "months") # generate sequence of months between Jan 2010 and Jan 2012
months
```

### Factor 

- **Factor** is a special type of character variable, often used to indicate membership in one of several possible categories, called **levels** (e.g. for plotting, or conditional statistics and data work)
```{r factors}
students<-factor(c("freshman", "senior", "senior", "junior", "freshman", 
                   "sophomore", "freshman"))
students # note order is arbitrary
levels(students) #extract unique levels 
nlevels(students) #count the number of levels 
```

```{r factors2}
table(students) #tabulate number of values for each level 
```

#### Ordered Factors

- Factors have ordered `levels()` which control the order on plots and in `table()` 

```{r}
students.o<-ordered(students, levels=c("freshman","sophomore","junior","senior"))
students.o
```

- **Be advised**: when `R` stores and calls factors, it actually stores them as integers [1..k, for k categories] instead of characters (e.g. "freshman"=1, "sophomore"=2), making this a **nominal** variable. This allows for some mathematical operations.
- An **ordered factor** is where the ordering matters (e.g. "small", "medium", "large" coded as 1, 2, 3 in order)

## Checking or Reclassifying Objects

- We can always check the class of an object with `class()` or `typeof()`. 
    - We can perform logical tests `is.numeric()`, `is.factor()`, etc. to see if an object is a specified type
    - We can change the class of an object by redefining it with `as.classname()`, e.g.

```{r class change}
x<-1:5
is.numeric(x) # check if x is numeric
is.factor(x) # check if x is a character
x<-as.character(x) # change vector x to a character
class(x)
x<-as.numeric(x) # change vector x back to numeric 
class(x)
```
